---
title: "shiny app graphs"
output: github_document
date: "2025-05-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(shiny)
library(tidyverse)
library(readxl)
library(openintro)
library(ggplot2)
```

## Getting the data

```{r}
college_aid <- read_excel("college aid 2000-2022.xlsx", skip = 2)
college_aid <- college_aid |>
  rename(Year = `Control and level of institution, and year`)

college_aid <- college_aid |>
  mutate(Type = ifelse(is.na(`Number enrolled`), Year, NA)) |>
  fill(Type) |>
  mutate(Type = str_squish(Type), 
         Year = str_squish(Year)) |>
  filter(!is.na(`Number enrolled`)) |>
  mutate(Level = ifelse(Type %in% c("2-year", "4-year"), Type, "Both")) |>
  select(Type, Level, everything()) |>
  mutate(Type = ifelse(Type %in% c("2-year", "4-year"), NA, Type)) |>
  fill(Type)


names(college_aid)<- c(
  "type",
  "level",
  "year",
  "num_of_enrolled",
  "num_awarded_aid",
  "pct_sutdents_awarded_aid",
  "pct_with_fed_grants", 
  "pct_with_state_local_grants", 
  "pct_with_institutional_grant", 
  "pct_student_loans",
  "avg_federal_grants_award", 
  "avg_state_local_grants_award", 
  "avg_inst_grants_award", 
  "avg_student_loans_award",
  "avg_federal_grants_const", 
  "avg_state_local_grants_const", 
  "avg_inst_grants_const", 
  "avg_student_loans_const"
)

college_aid <- college_aid |>
  mutate(across(
    .cols = num_of_enrolled:avg_student_loans_const,
    .fns  = ~ round(as.numeric(.), 3)
  ))

aid_group_long <- college_aid |>
  pivot_longer(
    cols = c(avg_federal_grants_award, 
             avg_state_local_grants_award, 
             avg_inst_grants_award, 
             avg_student_loans_award),
    values_to = "avg_award",
    names_to = "aid_type") |>
  mutate(
    aid_type = recode(aid_type,
                      avg_federal_grants_award   = "Federal Grants",
                      avg_state_local_grants_award = "State/Local Grants",
                      avg_inst_grants_award = "Institutional Grants",
                      avg_student_loans_award = "Student Loans"),
    year = factor(year, levels = unique(year))) |>
  select(type, level, year, aid_type, avg_award)

#still need to remove the first two rows
college_aid <- college_aid[-c(1,2), ]
```


## stats for private non-profit vs private for-profit

```{r}
#difference between private non-profit and private for-profit
college_aid_subset1 <- college_aid[college_aid$type %in% c("Private nonprofit", "Private for-profit"), ]

college_aid_subset_nonvsfor <- college_aid_subset1[college_aid_subset1$level %in% c("Both"), ]


```



## Public vs private non-profit
```{r}
#difference between private non-profit and public for average student loans award
college_aid_avg_stud_loan <- college_aid[college_aid$type %in% c("Private nonprofit", "Public"), ]

college_aid_subset_privvspub <- college_aid_avg_stud_loan[college_aid_avg_stud_loan$level %in% c("Both"), ]


```


### Linear models?

```{r}
all_inst_subset <- college_aid[college_aid$type %in% c("All institutions"), ]

all_inst_subset$year_numeric <- as.numeric(sub("-.*", "", all_inst_subset$year))

all_inst_subset$avg_student_loans_award <- as.numeric(all_inst_subset$avg_student_loans_award)

all_institutions_loans_model <- lm(avg_student_loans_award~year_numeric, data = all_inst_subset)

summary(all_institutions_loans_model)
```


```{r}
ggplot(all_inst_subset, aes(x = year_numeric, y = avg_student_loans_award)) + 
  geom_point() +
  geom_smooth(method= 'lm') 
```

### trying to turn the data linear
The model below shows public vs private non-profit and it shows that the type was non-significant, and so was the interaction between type and year, however, year was significant which means that average student loan awards have change over time but that change is not dependent on type of university. so they do show a trend on increasing, but the trend is similar between public and private and their is no significant difference between the two.

```{r}
#first for private nonprofit vs public
college_aid_subset_privvspub$year_numeric <- as.numeric(sub("-.*", "", college_aid_subset_privvspub$year))

public_subset <- college_aid_subset_privvspub[college_aid_subset_privvspub$type %in% c("Public"), ]

private_nonprofit_subset <- college_aid_subset_privvspub[college_aid_subset_privvspub$type %in% c("Private nonprofit"), ]

#look at plot for public (looks way more linear)
ggplot(public_subset, aes(x = year_numeric, y = avg_student_loans_award)) + 
  geom_point() +
  geom_smooth(method= 'lm')

#look at plot for private nonprofit (also more linear)
ggplot(private_nonprofit_subset, aes(x = year_numeric, y = avg_student_loans_award)) + 
  geom_point() +
  geom_smooth(method= 'lm')


#model with interaction
pubvpriv_model <- lm(avg_student_loans_award ~ type + year_numeric + type:year_numeric, data = college_aid_subset_privvspub)

#model without interaction
pubvpriv_model1 <- lm(avg_student_loans_award ~ type + year_numeric, data = college_aid_subset_privvspub)

summary(pubvpriv_model1)


#check assumption
plot(pubvpriv_model$fitted.values, resid(pubvpriv_model), main="Residuals vs. Fitted",
     xlab="Fitted values", ylab="Residuals")

qqnorm(resid(pubvpriv_model))

res1 <- resid(pubvpriv_model)
```

### Private non-profit vs private for-profit

This one does show significance for all three things which means that there is a significant difference between private non-profit and private for-profit, and that student loans do show a general trend that changes over time that is not the same rate for nonprofit and for-profit.

```{r}
college_aid_subset_nonvsfor$year_numeric <- as.numeric(sub("-.*", "", college_aid_subset_nonvsfor$year))

#try to see if private for profit is linear (not super linear)
private_forprofit_subset <- college_aid_subset_nonvsfor[college_aid_subset_nonvsfor$type %in% c("Private for-profit"), ]

ggplot(private_forprofit_subset, aes(x = year_numeric, y = avg_student_loans_award)) + 
  geom_point() +
  geom_smooth(method= 'lm')

#try a model
nonprofitvforprofit_model <- lm(avg_student_loans_award ~ type + year_numeric + type:year_numeric, data = college_aid_subset_nonvsfor)

summary(nonprofitvforprofit_model)

#check assumptions
res <- resid(nonprofitvforprofit_model)

plot(nonprofitvforprofit_model$fitted.values, resid(nonprofitvforprofit_model), main="Residuals vs. Fitted",
     xlab="Fitted values", ylab="Residuals")

```

```{r}
ggplot(college_aid_subset_nonvsfor, aes(x = year_numeric, y = avg_student_loans_award, color = type)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Interaction Between Year & University Type",
       x = "Year", y = "Avg Student Loans Award")

ggplot(college_aid_subset_privvspub, aes(x = year_numeric, y = avg_student_loans_award, color = type)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Interaction Between Year & University Type",
       x = "Year", y = "Avg Student Loans Award")
```

### Public vs private for-profit
i am unsure if I should keep the transformation here, if it really does anything

```{r}
college_aid_subset_pubvprivforprofit <- college_aid[college_aid$type %in% c("Public", "Private for-profit"), ]

college_aid_subset_pubvprivforprofit1 <- college_aid_subset_pubvprivforprofit[college_aid_subset_pubvprivforprofit$level %in% c("Both"), ]

college_aid_subset_pubvprivforprofit1$year_numeric <- as.numeric(sub("-.*", "", college_aid_subset_pubvprivforprofit1$year))

#model
pubvprivateforprofit_model <- lm(avg_student_loans_award ~ type + year_numeric + type:year_numeric, data = college_aid_subset_pubvprivforprofit1)

summary(pubvprivateforprofit_model)

#check assumptions
res2 <- resid(pubvprivateforprofit_model)
qqnorm(res2)
qqline(res2)

plot(pubvprivateforprofit_model$fitted.values, resid(pubvprivateforprofit_model), main="Residuals vs. Fitted",
     xlab="Fitted values", ylab="Residuals")

#look at a plot
ggplot(college_aid_subset_pubvprivforprofit1, aes(x = year_numeric, y = avg_student_loans_award, color = type)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Interaction Between Year & University Type",
       x = "Year", y = "Avg Student Loans Award")

# attempting a log transformation
college_aid_subset_pubvprivforprofit1$log_y <- log(college_aid_subset_pubvprivforprofit1$avg_student_loans_award)

log_pubvforprofit_model <- lm(log_y ~ type * year_numeric, data = college_aid_subset_pubvprivforprofit1)

summary(log_pubvforprofit_model)

#re-check assumptions with log
reslog <- resid(log_pubvforprofit_model)
qqnorm(reslog)
qqline(reslog)

#look at a plot for log
ggplot(college_aid_subset_pubvprivforprofit1, aes(x = year_numeric, y = log_y, color = type)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Interaction Between Year & University Type",
       x = "Year", y = "Avg Student Loans Award")
```


### Regression for institutional grants

```{r}
#public vs private nonprofit

#model with interaction
pubvpriv_model_inst_grants <- lm(avg_inst_grants_award ~ type + year_numeric + type:year_numeric, data = college_aid_subset_privvspub)

summary(pubvpriv_model_inst_grants)


#public vs private for profit
#model with no interaction
pubvprivateforprofit_model_inst_grants <- lm(avg_inst_grants_award ~ type + year_numeric, data = college_aid_subset_pubvprivforprofit1)

summary(pubvprivateforprofit_model_inst_grants)


#private vs private
nonprofitvforprofit_model_inst_grants <- lm(avg_inst_grants_award ~ type + year_numeric + type:year_numeric, data = college_aid_subset_nonvsfor)

summary(nonprofitvforprofit_model_inst_grants)
```

###Regression for federal grants

```{r}
#public vs private nonprofit

#model with interaction (non)
pubvpriv_model_fed_grants <- lm(avg_federal_grants_award ~ type + year_numeric + type:year_numeric, data = college_aid_subset_privvspub)

summary(pubvpriv_model_fed_grants)


#public vs private for profit
#model with no interaction
pubvprivateforprofit_model_fed_grants <- lm(avg_federal_grants_award ~ type + year_numeric, data = college_aid_subset_pubvprivforprofit1)

summary(pubvprivateforprofit_model_fed_grants)


#private vs private
nonprofitvforprofit_model_fed_grants <- lm(avg_federal_grants_award ~ type + year_numeric + type:year_numeric, data = college_aid_subset_nonvsfor)

summary(nonprofitvforprofit_model_fed_grants)
```

###Regression for state/local
```{r}
#public vs private nonprofit

#model with interaction (non)
pubvpriv_model_state_grants <- lm(avg_state_local_grants_award ~ type + year_numeric + type:year_numeric, data = college_aid_subset_privvspub)

summary(pubvpriv_model_state_grants)


#public vs private for profit
#model with no interaction
pubvprivateforprofit_model_state_grants <- lm(avg_state_local_grants_award ~ type + year_numeric + type:year_numeric, data = college_aid_subset_pubvprivforprofit1)

summary(pubvprivateforprofit_model_state_grants)


#private vs private
nonprofitvforprofit_model_state_grants <- lm(avg_state_local_grants_award ~ type + year_numeric + type:year_numeric, data = college_aid_subset_nonvsfor)

summary(nonprofitvforprofit_model_state_grants)
```

